---
title: "Case Study - Bike Share: Data Cleaning & Transformation"
author: "Jannatul Ashpia"
output: html_document
---
```{r warning=FALSE}
#Load libraries
library(tidyverse)
library(lubridate)
library(ggplot2)
```


```{r echo=FALSE, warning=FALSE, show_col_types=FALSE}
#read data frames
april_Data <- read_csv('202304-divvy-tripdata.csv', show_col_types=FALSE)
may_Data <- read_csv('202305-divvy-tripdata.csv', show_col_types=FALSE)
june_Data <- read_csv('202306-divvy-tripdata.csv', show_col_types=FALSE)
```

```{r error=FALSE, message=FALSE}
glimpse(april_Data)
```
```{r}
glimpse(may_Data)
```
```{r}
glimpse(june_Data)
```
**Merge the 3 data frames**

```{r}
merged_Df <- bind_rows(april_Data, may_Data, june_Data)
```
## Clean up
**Information Regarding longitude and Latitude is not necessary for our case, thus it can be removed**
```{r}
ride_data <- merged_Df %>% 
  select(1:8,13) 

colnames(ride_data)
summary(ride_data)
```
#### Changes to apply to the Dataset
* Add a _ride_length_ column calculating the duration for trips
* Remove rows for column _ride_length_ if it is less than 0
* Split the _started_at_ column further into _month_, _day_, _days of week_

```{r}
#add ride_length column
ride_data <- ride_data %>% 
  mutate(ride_length = ended_at - started_at)
```
This results in ride_length to appear as **seconds**
```{r}
#convert it to numeric
ride_data$ride_length <- as.numeric(as.character(ride_data$ride_length))
```

```{r}
#check for negative values
ride_data%>%
  select(ride_length) %>% 
  filter(ride_length < 0)
```


```{r}
#add columns for month, date, day
ride_data <- ride_data %>% 
  mutate(month = format(as.Date(started_at), "%m"),
         day = format(as.Date(started_at), "%d"),
         day_of_week = format(as.Date(started_at), "%A"),
         hour = hour(started_at))

# create a new data frame excluding negative ride_length
trip_data <- ride_data[!(ride_data$ride_length<0),]
```

## Descriptive Analysis
**Note**: ride_length is calculated in _seconds_
```{r}
#analyzing ride_length
summary(trip_data$ride_length)
```

#### Casual vs Member Statistics on ride_length
```{r}
trip_data %>% 
  group_by(member_casual) %>% 
  summarise(avg = mean(ride_length),
            min = min(ride_length),
            median = median(ride_length),
            max = max(ride_length))
```

```{r include=FALSE}
#fix weekdays ordering
trip_data$day_of_week <- ordered(trip_data$day_of_week, 
                                    levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

#### Weekdays ride_length comparision for the member types

```{r warning=FALSE, message=FALSE}
trip_data %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(avg = mean(ride_length),
            min = min(ride_length),
            median = median(ride_length),
            max = max(ride_length)) %>% 
  arrange(member_casual, day_of_week)
```

#### Total weekdays ride_length among member types
```{r warning=FALSE, message=FALSE}
trip_data %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(total_ride = n(),
            avg_duration = mean(ride_length)) %>% 
  arrange(member_casual, day_of_week)
```

```{r eval=FALSE, include=FALSE}
write_csv(trip_data, file = "/Volumes/Seagate_Por/GoogleAnalytics/csv/trip_data.csv")
```

